---
title: "Processing Code"
author: "Ranni Tewfik"
date: "03/12/2024"
output: html_document
---


# Processing Script


This processing script loads the raw data, processes and cleans it, and saves it as an Rds file in the processed-data folder.


# Load and Check Raw Data


```{r}
# Load required packages
library(here)
library(dplyr)

# Load raw data
seer1 <- read.csv(here("data/raw-data/SEER.csv"))

# Summary of raw data
glimpse(seer1)
summary(seer1)
head(seer1)
```


The raw dataset has 1,048,575 observations and 7 variables: "race", "age", "stage", "survival_months", "vital_status", "education", and "family_history". All variables are character class.


# Explore Raw Data


```{r}
# Attach raw data
attach(seer1)

# Get frequencies of categorical variable "race"
table(race, useNA = "always")

# Get frequencies of categorical variable "age" 
table(age, useNA = "always")

# Get frequencies of categorical variable "stage" 
table(stage, useNA = "always")

# Get frequencies of categorical variable "vital_status" 
table(vital_status, useNA = "always")

# Get frequencies of categorical variable "education" 
table(education, useNA = "always")

# Get frequencies of categorical variable "family_history" 
table(family_history, useNA = "always")

# Convert "survival_months" variable into numeric class and get summary of the variable
survival_months <- as.numeric(survival_months)
summary(survival_months)
```


There are six categories for the "race" variable: "Hispanic (All Races)", "Non-Hispanic White", "Non-Hispanic Black", "Non-Hispanic Asian or Pacific Islander", "Non-Hispanic American Indian/Alaska Native", and "Non-Hispanic Unknown Race." There are no missing data for the "race" variable.

There are 19 categories for the "age" variable: "00 years", "01-04 years", "05-09 years", "10-14 years", "15-19 years", "20-24 years", "25-29 years", "30-34 years", "35-39 years", "40-44 years", "45-49 years", "50-54 years", "55-59 years", "60-64 years", "65-69 years", "70-74 years", "75-79 years", "80-84 years", and "85+ years". There are no missing data for the "age" variable. 

There are six categories for the "stage" variable: "Blank(s)", "Distant", "In situ", "Localized", "Regional", and "Unknown/unstaged". There are no missing data for the "stage" variable.

There are two categories for the "vital_status" variable: "Alive" and "Dead". There are no missing data for the "vital_status" variable.

There are three categories for the "education" variable: "high school or less", "some college", and "college or higher". There are no missing data for the "education" variable.

There are two categories for the "family_history" variable: "yes" and "no". There are no missing data for the "family_history" variable.

For the "survival_months" variable, the mean is 88.46 months, the minimum is 0 months, and the maximum is 227 months. There are 8,217 observations with missing data for the "survival_months" variable.


# Clean Raw Data


```{r}
# Create a new dataset from the raw data that does not include observations with missing data for "survival_months"; includes only "Localized", "Regional", and "Distant" categories for "stage"; and creates a new "death" variable from "vital_status" in which "death" = 1 if "vital_status" = "Dead" and "death" = 0 if "vital_status" = "Alive"
seer2 <- seer1 %>% 
  filter((survival_months != "Unknown") & (stage == "Localized" | stage == "Regional" | stage == "Distant")) %>%
  mutate(death = ifelse(vital_status == "Dead", 1, 0))

# Create a new variable "ethnicity" in the new dataset that uses the categories in the "race" variable and groups them into five new categories: "White", "Black", "Asian", "Hispanic", and "Other"
seer2$ethnicity <- ifelse(seer2$race == "Non-Hispanic White", "White", 
                         ifelse(seer2$race == "Non-Hispanic Black", "Black", 
                                ifelse(seer2$race == "Non-Hispanic Asian or Pacific Islander", "Asian", 
                                       ifelse(seer2$race == "Hispanic (All Races)", "Hispanic", 
                                              ifelse(seer2$race == "Non-Hispanic American Indian/Alaska Native" | seer2$race == "Non-Hispanic Unknown Race", "Other", NA)))))

# Create a new variable "agegroup" in the new dataset that uses the categories in the "age" variable and groups them into five new categories: "0-49 years", "50-59 years", "60-69 years", "70-79 years", and "80+ years"
seer2$agegroup <- ifelse(is.na(seer2$age), NA,
                         ifelse(seer2$age == "50-54 years" | seer2$age == "55-59 years", "50-59 years",
                                ifelse(seer2$age == "60-64 years" | seer2$age == "65-69 years", "60-69 years",
                                       ifelse(seer2$age == "70-74 years" | seer2$age == "75-79 years", "70-79 years",
                                              ifelse(seer2$age == "80-84 years" | seer2$age == "85+ years", "80+ years", "0-49 years")))))

# Create a new numeric variable "survival_time" in the new dataset from the variable "survival_months"
seer2$survival_time <- seer2$survival_months
seer2$survival_time <- as.numeric(seer2$survival_time)

# Convert the variables "ethnicity", "agegroup", "stage", "death", "education", and "family_history" to factor class
seer2$ethnicity <- as.factor(seer2$ethnicity)
seer2$agegroup <- as.factor(seer2$agegroup)
seer2$stage <- as.factor(seer2$stage)
seer2$death <- as.factor(seer2$death)
seer2$education <- as.factor(seer2$education)
seer2$family_history <- as.factor(seer2$family_history)

# Keep only the following variables in the new dataset: "ethnicity", "agegroup", "stage", "education", "family_history", "survival_time",  and "death"
seer <- subset(seer2, select = c(ethnicity, agegroup, stage, education, family_history, survival_time, death))

# Summary of processed data
glimpse(seer)
summary(seer)
head(seer)
```


After processing the raw dataset to remove missing observations for "survival_months" (represented as "survival_time" in the processed dataset) and only including the "Localized", "Regional", and "Distant" categories for the "stage" variable, the processed dataset has 776,196 observations and 5 variables: "ethnicity", "agegroup", "stage", "survival_time", and "death". The variable "survival_time" is numeric class, the variable "death" is factor class, and the remaining variables are character class.


# Save Processed Data


```{r}
saveRDS(seer, file = here("data/processed-data/processeddata.rds"))
```


The processed data ("processeddata.rds") can be found in the "processed-data" folder.

