---
title: "Exploratory Analysis"
author: "Ranni Tewfik"
date: "02/23/2024"
output: html_document
---

  
# Exploratory Analysis

This exploratory analysis script loads the processed data and explores the five variables in the dataset.


# Load and Check Processed Data

```{r}
# Load required packages
library(here)
library(dplyr)
library(skimr)
library(survival)
library(survminer)
library(ggplot2)
library(tidyverse)
library(gtsummary)

# Load processed data
seer <- readRDS(here("data/processed-data/processeddata.rds"))

# Attach processed data
attach(seer)

# Summary of processed data
glimpse(seer)
summary(seer)
head(seer)
```


# Data Exploration - Tables

```{r}
# Frequency and proportion tables of "ethnicity" variable
table(ethnicity, useNA = "always")
prop.table(table(ethnicity))

# Frequency and proportion tables of "agegroup" variable
table(agegroup, useNA = "always")
prop.table(table(agegroup))

# Frequency and proportion tables of "stage" variable
table(stage, useNA = "always")
prop.table(table(stage))

# Frequency and proportion tables of "death" variable
table(death, useNA = "always")
prop.table(table(death))

# Summary of "survival_time" variable
summary(survival_time)

# Create a summary table of the data
summary_seer <- skim(seer)
print(summary_seer)

# Save the summary table
saveRDS(summary_seer, file = here("results/tables/summarytable.rds"))
```


A summary table of the data ("summarytable.rds") can be found in the "tables" folder.


# Table 1 - Study sample characteristics

```{r}
# Create Table 1 (study sample characteristics) from the data
table1 <- seer %>% tbl_summary(by = ethnicity, 
                               statistic = list(all_continuous() ~ "{mean} ({sd})"),
                               label = c(agegroup ~ "Age at diagnosis (years)", stage ~ "Cancer stage at diagnosis",      
                               survival_time ~ "Survival time (months)", death ~ "Vital status")) %>%
                   modify_caption("Table 1. Study sample characteristics of males in the U.S. with prostate cancer by race                                   and ethnicity.")

# Save the summary table
saveRDS(table1, file = here("results/tables/table1.rds"))
```


Table 1 of the study sample characteristics ("table1.rds") can be found in the "tables" folder.


# Table 2 - Kaplan-Meier survival
```{r}
# Create Table 2 (Kaplan-Meier 1-year, 5-year, and 10-year survival from the data)
## Overall survival
summary(survfit(Surv(survival_time, death) ~ 1), times <- c(12, 60, 120))

## Survival by "ethnicity"
summary(survfit(Surv(survival_time, death) ~ ethnicity), times <- c(12, 60, 120))
survdiff(Surv(survival_time, death) ~ ethnicity)

## Survival by "agegroup"
summary(survfit(Surv(survival_time, death) ~ agegroup), times <- c(12, 60, 120))
survdiff(Surv(survival_time, death) ~ agegroup)

## Survival by "stage"
summary(survfit(Surv(survival_time, death) ~ stage), times <- c(12, 60, 120))
survdiff(Surv(survival_time, death) ~ stage)
```


# Data Exploration - Figures

```{r}
survtime_hist <- seer %>% ggplot(aes(x = survival_time)) + geom_histogram() 
plot(survtime_hist)

ggsave(filename = here("results/figures/survtime_distribution.png"), plot = survtime_hist) 
```

A histogram of the distribution of the "survival_time" variable ("survtime_distribution.png") can be found in the "figures" folder.


```{r}
# Create a Kaplan-Meier survival curve stratified by ethnicity
plot(survfit(Surv(survival_time, death) ~ ethnicity), 
     col = c("red", "blue", "limegreen", "orange", "magenta"),
     xlab = "Survival Time (Months)",
     ylab = "Survival Probability")

legend(x = "topleft", legend = c("Non-Hispanic White", "Non-Hispanic Black", "Non-Hispanic Asian or Pacific Islander", "Hispanic (All Races)", "Other"),
       lty = c(1, 1, 1, 1, 1),
       col = c("magenta", "blue", "red", "limegreen", "orange"))

title("Kaplan-Meier Survival Estimates")

# Save the Kaplan-Meier survival curve 
png(here("results/figures/survivalcurve.png"))

plot(survfit(Surv(survival_time, death) ~ ethnicity), 
     col = c("red", "blue", "limegreen", "orange", "magenta"),
     xlab = "Survival Time (Months)",
     ylab = "Survival Probability")

legend(x = "topleft", legend = c("Non-Hispanic White", "Non-Hispanic Black", "Non-Hispanic Asian or Pacific Islander", "Hispanic (All Races)", "Other"),
       lty = c(1, 1, 1, 1, 1),
       col = c("magenta", "blue", "red", "limegreen", "orange"))

title("Kaplan-Meier Survival Estimates")

dev.off()
```


A Kaplan-Meier survival curve stratified by ethnicity ("survivalcurve.png") can be found in the "figures" folder.

