---
title: "Exploratory Analysis"
author: "Ranni Tewfik"
date: "03/12/2024"
output: html_document
---

  
# Exploratory Analysis


This exploratory analysis script loads the processed data and explores the 7 variables in the dataset.


# Load and Check Processed Data


```{r}
# Load required packages
library(here)
library(dplyr)
library(skimr)
library(survival)
library(survminer)
library(ggplot2)
library(tidyverse)
library(gtsummary)
library(car)
library(gt)
library(webshot2)

# Load processed data
seer <- readRDS(here("data/processed-data/processeddata.rds"))

# Attach processed data
attach(seer)

# Summary of processed data
glimpse(seer)
summary(seer)
head(seer)
```


# Data Exploration - Tables


```{r}
# Frequency and proportion tables of "ethnicity" variable
table(ethnicity, useNA = "always")
prop.table(table(ethnicity))

# Frequency and proportion tables of "agegroup" variable
table(agegroup, useNA = "always")
prop.table(table(agegroup))

# Frequency and proportion tables of "stage" variable
table(stage, useNA = "always")
prop.table(table(stage))

# Frequency and proportion tables of "education" variable
table(education, useNA = "always")
prop.table(table(education))

# Frequency and proportion tables of "family_history" variable
table(family_history, useNA = "always")
prop.table(table(family_history))

# Frequency and proportion tables of "death" variable
table(death, useNA = "always")
prop.table(table(death))

# Summary of "survival_time" variable
summary(survival_time)

# Create a summary table of the data
summary_seer <- skim(seer)
print(summary_seer)

# Save the summary table
saveRDS(summary_seer, file = here("results/tables/summarytable.rds"))
```


A summary table of the data ("summarytable.rds") can be found in the "tables" folder.


# Table 1 - Study sample characteristics


```{r}
# Create Table 1 (study sample characteristics) from the data
table <- seer %>% 
  mutate(death = case_when(death == "1" ~ "alive",
                           death == "2" ~ "dead")) %>%
  tbl_summary(by = ethnicity, 
              statistic = list(all_continuous() ~ "{mean} ({sd})"),
              label = c(agegroup ~ "Age at diagnosis (years)", stage ~ "Cancer stage at diagnosis",   
                        education ~ "Educational attainment", family_history ~ "Family history of cancer",
                        survival_time ~ "Survival time (months)", death ~ "Vital status")) %>%
  modify_caption("Table 1. Study sample characteristics of males in the U.S. with prostate cancer by race and ethnicity.")

table1 <- as_gt(table)

table1

# Save the summary table
gtsave(table1, file = here("results/tables/table1.png"))
```


Table 1 of the study sample characteristics ("table1.rds") can be found in the "tables" folder.


Kaplan-Meier Survival Tables


```{r}
# Look at Kaplan-Meier 1-year, 5-year, and 10-year survival from the data
## Overall survival
summary(survfit(Surv(survival_time, death) ~ 1), times <- c(12, 60, 120))

## Survival by "ethnicity"
summary(survfit(Surv(survival_time, death) ~ ethnicity), times <- c(12, 60, 120))

## Survival by "agegroup"
summary(survfit(Surv(survival_time, death) ~ agegroup), times <- c(12, 60, 120))

## Survival by "stage"
summary(survfit(Surv(survival_time, death) ~ stage), times <- c(12, 60, 120))
```


# Data Exploration - Figures


```{r}
# Create a histogram of "survival_time"
survtime_hist <- seer %>% ggplot(aes(x = survival_time)) + geom_histogram() 
plot(survtime_hist)

# Save the histogram
ggsave(filename = here("results/figures/survtime_distribution.png"), plot = survtime_hist) 
```

A histogram of the distribution of the "survival_time" variable ("survtime_distribution.png") can be found in the "figures" folder.


```{r}
# Create a boxplot of "death" and "survival_time"
ggplot(seer, aes(x = factor(death), y = survival_time)) +
  geom_boxplot() +
  labs(x = "Death (no vs. yes)", y = "Survival Time (months)", title = "Boxplot of Mortality and Survival Time")

# Create a boxplot of "ethnicity" and "survival_time"
ggplot(seer, aes(x = factor(ethnicity), y = survival_time)) +
  geom_boxplot() +
  labs(x = "Ethnicity", y = "Survival Time (months)", title = "Boxplot of Ethnicity and Survival Time")

# Create a boxplot of "agegroup" and "survival_time"
ggplot(seer, aes(x = factor(agegroup), y = survival_time)) +
  geom_boxplot() +
  labs(x = "Age Group", y = "Survival Time (months)", title = "Boxplot of Age Groups and Survival Time")

# Create a boxplot of "stage" and "survival_time"
ggplot(seer, aes(x = factor(stage), y = survival_time)) +
  geom_boxplot() +
  labs(x = "Cancer Stage", y = "Survival Time (months)", title = "Boxplot of Cancer Stage and Survival Time")

# Create a boxplot of "education" and "survival_time"
ggplot(seer, aes(x = factor(education), y = survival_time)) +
  geom_boxplot() +
  labs(x = "Education", y = "Survival Time (months)", title = "Boxplot of Education and Survival Time")

# Create a boxplot of "family_history" and "survival_time"
ggplot(seer, aes(x = factor(family_history), y = survival_time)) +
  geom_boxplot() +
  labs(x = "Family History", y = "Survival Time (months)", title = "Boxplot of Family History and Survival Time")
```


```{r}
# Create a boxplot of "death" and "survival_time", stratified by "ethnicity"
death_survtime_ethnicity_box <- seer %>% ggplot(aes(x = factor(death), y = survival_time, fill = ethnicity)) +
  geom_boxplot() +
  labs(x = "Death (no vs. yes)", y = "Survival Time (months)", title = "Boxplot of Mortality and Survival Time, Stratified by Ethnicity")

# Save the boxplot
ggsave(filename = here("results/figures/death_survtime_ethnicity_boxplot.png"), plot = death_survtime_ethnicity_box) 
```


Boxplots of the distribution of the "survival_time" variable for "death" (no vs. yes), stratified by "ethnicity" ("death_survtime_ethnicity_boxplot.png") can be found in the "figures" folder.


```{r}
# Create a bar chart of "ethnicity" and "survival_time", stratified by "death"
ethnicity_survtime_death_bar <- seer %>% ggplot(aes(x = factor(ethnicity), y = survival_time, fill = death)) +
  geom_bar(position = "dodge", stat = "identity") +
  labs(x = "Ethnicity", y = "Survival Time (months)", fill = "Death (no vs. yes)", title = "Bar Chart of Ethnicity and Survival Time, Stratified by Death")

# Save the bar chart
ggsave(filename = here("results/figures/ethnicity_survtime_death_barchart.png"), plot = ethnicity_survtime_death_bar) 
```


Bar charts of the frequency of the "survival_time" variable for "ethnicity", stratified by "death" (no vs. yes) ("ethnicity_survtime_death_barchart.png") can be found in the "figures" folder.


```{r}
# Create residual plots
model <- lm(survival_time ~ death + ethnicity + agegroup + stage + education + family_history, data = seer)
residualPlots(model)
```


```{r}
# Create an overall Kaplan-Meier survival curve
seer$death <- as.numeric(seer$death)

fit <- survfit(Surv(survival_time, death) ~ 1, data = seer)

survplot <- ggsurvplot(fit, data = seer, 
                       censor.size = 0.1, 
                       size = 0.2, 
                       ylim = c(0.50, 1.00)) +
  labs(x = "Survival Time (months)", y = "Survival Probability", title = "Kaplan-Meier Survival Estimates")

# Save the Kaplan-Meier survival curve
ggsave(filename = here("results/figures/survivalcurve_total.png"), plot = survplot$plot) 
```


An overall Kaplan-Meier survival curve ("survivalcurve_total.png") can be found in the "figures" folder.


```{r}
# Create a Kaplan-Meier survival curve stratified by "ethnicity"
seer$death <- as.numeric(seer$death)

fit <- survfit(Surv(survival_time, death) ~ ethnicity, data = seer)

survplot <- ggsurvplot(fit, data = seer, 
                       censor.size = 0.1, 
                       size = 0.2, 
                       ylim = c(0.50, 1.00), 
                       legend.labs = c("Asian or Pacific Islander", "Black", "Hispanic", "Other", "White")) +
  labs(x = "Survival Time (months)", y = "Survival Probability", title = "Kaplan-Meier Survival Estimates, Stratified by Ethnicity")

# Save the Kaplan-Meier survival curve
ggsave(filename = here("results/figures/survivalcurve_ethnicity.png"), plot = survplot$plot) 
```


A Kaplan-Meier survival curve stratified by "ethnicity" ("survivalcurve_ethnicity.png") can be found in the "figures" folder.


```{r}
# Create a Kaplan-Meier survival curve stratified by "agegroup"
seer$death <- as.numeric(seer$death)

fit <- survfit(Surv(survival_time, death) ~ agegroup, data = seer)

survplot <- ggsurvplot(fit, data = seer, 
                       censor.size = 0.1, 
                       size = 0.2, 
                       ylim = c(0.50, 1.00)) +
  labs(x = "Survival Time (months)", y = "Survival Probability", title = "Kaplan-Meier Survival Estimates, Stratified by Age")

# Save the Kaplan-Meier survival curve
ggsave(filename = here("results/figures/survivalcurve_age.png"), plot = survplot$plot) 
```


A Kaplan-Meier survival curve stratified by "agegroup" ("survivalcurve_age.png") can be found in the "figures" folder.


```{r}
# Create a Kaplan-Meier survival curve stratified by "stage"
seer$death <- as.numeric(seer$death)

fit <- survfit(Surv(survival_time, death) ~ stage, data = seer)

survplot <- ggsurvplot(fit, data = seer, 
                       censor.size = 0.1, 
                       size = 0.2, 
                       ylim = c(0.50, 1.00)) +
  labs(x = "Survival Time (months)", y = "Survival Probability", title = "Kaplan-Meier Survival Estimates, Stratified by Stage")

# Save the Kaplan-Meier survival curve
ggsave(filename = here("results/figures/survivalcurve_stage.png"), plot = survplot$plot) 
```


A Kaplan-Meier survival curve stratified by "stage" ("survivalcurve_stage.png") can be found in the "figures" folder.


```{r}
# Create a Kaplan-Meier survival curve stratified by "education"
seer$death <- as.numeric(seer$death)

fit <- survfit(Surv(survival_time, death) ~ education, data = seer)

survplot <- ggsurvplot(fit, data = seer, 
                       censor.size = 0.1, 
                       size = 0.2, 
                       ylim = c(0.50, 1.00)) +
  labs(x = "Survival Time (months)", y = "Survival Probability", title = "Kaplan-Meier Survival Estimates, Stratified by Education")

# Save the Kaplan-Meier survival curve
ggsave(filename = here("results/figures/survivalcurve_education.png"), plot = survplot$plot) 
```


A Kaplan-Meier survival curve stratified by "education" ("survivalcurve_education.png") can be found in the "figures" folder.


```{r}
# Create a Kaplan-Meier survival curve stratified by "family_history"
seer$death <- as.numeric(seer$death)

fit <- survfit(Surv(survival_time, death) ~ family_history, data = seer)

survplot <- ggsurvplot(fit, data = seer, 
                       censor.size = 0.1, 
                       size = 0.2, 
                       ylim = c(0.50, 1.00)) +
  labs(x = "Survival Time (months)", y = "Survival Probability", title = "Kaplan-Meier Survival Estimates, Stratified by Family History")

# Save the Kaplan-Meier survival curve
ggsave(filename = here("results/figures/survivalcurve_fhx.png"), plot = survplot$plot) 
```


A Kaplan-Meier survival curve stratified by "family_history" ("survivalcurve_fhx.png") can be found in the "figures" folder.

